A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-08-11 bioc-workflow.Rmd
Updated section on phenoDisco

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a)
[@hyper]. The data was generated using hyperplexed LOPIT (hyperLOPIT),
an improved method relying on improved sub-cellular fractionation and
more accurate quantitation, leading to more reliable classification of
protein localisation across the whole sub-cellular space. The method
uses an elaborate sub-cellular fractionation scheme, enabled by the
use of Tandem Mass Tag (TMT) [Thompson:2003] 10-plex and application
of the MS data acquisition technique named synchronous precursor
selection MS^3^ (SPS-MS^3^) [@McAlister:2014], for TMT
quantification with high accuracy and precision. Three biological
replicates were generated from the E14TG2a experiment, the first was
to target low density fractions and the second and third were to
emphasis separation of the denser organelles.  The intersect of
replicates 1 and 2 was treated as a 20-plex dataset for the analysis
discussed in the manuscript [@hyper] as it has been shown that
combining replicates across from different gradients can increase
spatial resolution [@Trotter:2010]. The combination of replicates resulted
in 5032 proteins common in both experiments.

@@

The extraction of sub-cellular protein clusters can be difficult owing
to the limited number of marker proteins that exist in databases and
elsewhere. Furthermore, given the vast complexity of the cell,
automatic annotation retrieval does not always give a full
representation of the true sub-cellular diversity in the data. For
downstream analyses, such as supervised machine learning, it is
desirable to obtain reliable markers that cover as many sub-cellular
niches as possible, as these markers are directly used in the training
phase of the ML classification. We find that a lack of sub-cellular
diversity in the labelled training data leads to prediction errors, as
unlabelled instances can only be assigned to a class that exists in
the training data [@Breckels:2013]. In such scenarios novelty
detection can be useful to identify data-specific sub-cellular
groupings such as organelles and protein complexes. The phenotype
discovery (phenoDisco) algorithm [@Breckels:2013] is one such method
and is available in `r Biocpkg("pRoloc")`. It is an iterative
semi-supervised learning method that combines the classification of
proteins on existing labelled data with the detection of new clusters.
Novelty detection methods are also useful for confirming the presence
of known clusters in an unbiased fashion. For example, in [@hyper] the
`phenoDisco` algorithm was used to detect the data-specific
confirmation and presence of the nucleus and nucleus sub-compartments,
as it was expected that nucleus associated clusters existed within the
data as the protocol included a separate chromatin-rich fraction to
enrich for nuclei. In the code chunk below, we show how to run a
series of novelty detection experiments using the `phenoDisco`
algorithm on the mouse stem-cell dataset. There are several optional
arguments that can be set dependent on the type of biological question
of interest. For example,

- Explain how to use different group sizes for detecting small complexes etc.
- ndims for using >=2 components
- times -> important for convergence
- input markers are different here -> explain choice of pd.markers

```{r pd, eval = FALSE}
## As per the hyperLOPIT paper
pdRes <- phenoDisco(hl, fcol = "phenoDisco.Input", times = 200, 
                    GS = 20, p = 0.05)

@@

>>>>

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a)
[@hyper]. The data was generated using hyperplexed LOPIT (hyperLOPIT),
a method relying on improved sub-cellular fractionation and
more accurate quantitation, leading to more reliable classification of
protein localisation across the whole sub-cellular space. The method
uses an elaborate sub-cellular fractionation scheme, enabled by the
use of Tandem Mass Tag (TMT) [Thompson:2003] 10-plex and application
of the MS data acquisition technique named synchronous precursor
selection MS^3^ (SPS-MS^3^) [@McAlister:2014], for TMT
quantification with high accuracy and precision. Three biological
replicates were generated from the E14TG2a experiment, the first was
to target low density fractions and the second and third were to
emphasis separation of the denser organelles.  The intersect of
replicates 1 and 2 was treated as a 20-plex dataset for the analysis
discussed in the manuscript [@hyper] as it has been shown that
combining replicates across from different gradients can increase
spatial resolution [@Trotter:2010]. The combination of replicates resulted
in 5032 proteins common in both experiments.

@@

The extraction of sub-cellular protein clusters can be difficult owing
to the limited number of marker proteins that exist in databases and
elsewhere. Furthermore, given the vast complexity of the cell,
automatic annotation retrieval does not always give a full
representation of the true sub-cellular diversity in the data. For
downstream analyses, such as supervised machine learning, it is
desirable to obtain reliable markers that cover as many sub-cellular
niches as possible, as these markers are directly used in the training
phase of the ML classification. We find that a lack of sub-cellular
diversity in the labelled training data leads to prediction errors, as
unlabelled instances can only be assigned to a class that exists in
the training data [@Breckels:2013]. In such scenarios novelty
detection can be useful to identify data-specific sub-cellular
groupings such as organelles and protein complexes. The phenotype
discovery (phenoDisco) algorithm [@Breckels:2013] is one such method
and is available in `r Biocpkg("pRoloc")`. It is an iterative
semi-supervised learning method that combines the classification of
proteins on existing labelled data with the detection of new clusters.

In addition to extracting new phenotypes, novelty detection methods 
are also useful for confirming the presence of known or postulated 
clusters in an unbiased fashion. For example, in [@hyper] 
the `phenoDisco` algorithm was used to confirm the data-specific
presence of the nucleus and nucleus sub-compartments. In the code 
chunk below, we re-run this analysis, highlighting some of the 
optional arguments and parameters available for phenotype extraction 
and give some advice on how to interpret the output.

As the `phenoDisco` algorithm is semi-supervised it uses both 
labelled (markers) and unlabelled data to explore the data structure 
and find new sub-cellular data clusters. Thus the first step is to 
define some input labelled data i.e. markers, that the algorithm 
will use as input for the supervised learning aspect of the 
algorithm. As described in [@hyper] we define a set of markers 
to use as input for the analyses that cover well-known residents 
from three distinct organelle structures; the mitochondria, plasma 
membrane and ER, and from three well-known and abundant protein 
complexes; the proteasome and two ribosomal subunits, 40S and 60S. 
These input markers are stored in the `featureData` column of 
`hl` where `fcol = "phenoDisoc.Input"`. We can use the convenience 
accessor function `getMarkers` to print out a table of the markers 
contained in this marker set. These initial markers were manually 
curated using information from the UniProt database, the Gene 
Ontology and the literature.

```{r displayPDmarkers}
getMarkers(hl, fcol = "phenoDisco.Input")
```

In the code chunk below we run the `phenoDisco` function and 
return a novelty detection result stored in `pdRes`, according 
to the specified parameters. The algorithm parameters `times`, 
`GS` and `p` are passed to the function, along with the `fcol` 
telling the algorithm where the input training data is contained.

```{r runPD, eval = FALSE}
## As per Christoforou et al (2016), 
pdRes <- phenoDisco(hl, fcol = "phenoDisco.Input", 
                    times = 200, 
                    GS = 60)

@@

The argument `times` defined the number of times we run unsupervied 
Gaussian Mixture Modelling before defining a new phenotype cluster. 
The recommended minimum and default value is 100. In the below 
code chunk we increase the value to `times = 200` as we have found 
for larger datasets (e.g. 5000+ proteins) a higher times is requried 
for convergence. `GS` defines the minimum number of proteins allowed 
per new data cluster and thus heavily influences what type of new 
clusters are extracted. For example, if a user is interesed in the 
detection of small complexes they may wish to use a small `GS = 10`, 
or `=20` etc. If they wish to detect larger more abundant sub-cellular 
niches a higher `GS` would be preferable. Specifying a small `GS` is be 
more time consuming than a larger `GS`, and there is a trade off 
between finding interesting small complexes and those that may not 
be of interest as we find there is a tendancy to find more noise when 
using a small `GS` compared to using a higher one.

One may also consider increasing the search space for new data 
clusters by increasing the value of the parameter `G`. This defines 
the number of GMM components to test and fit; the default is `G = 1:9` 
(`mclust`'s default). One should note that the choice of decreasing 
`GS`, and increasing the values of the arguments `times`, `G` 
(among other function arguments, see `?phenoDisco`) will heavily 
influence (increase) the total time taken to run the algorithm. 
`phenoDisco` supports parallelisation and we strongly suggest 
you make use of a parallel processing to run these analyses (different 
backends can be set with argument `BPPARAM`).

The ouput of running the `phenoDisco` algorithm is a `MSnSet` containing 
the new data clusters, appended to the `featureData` under the name 
`pd`. We can see by typing `pdRes` directly into the console the 
processing information has been updated to the `MSnSet` recording the 
parameters that were used to run the analyses. This is handy for keeping 
track of data analyses. The results can be displayed by using the 
`getMarkers` function. We see that 5 new phenotype data clusters were 
found.

```{r loadPDres, include=FALSE}
load("Data/pd_runs/pd.rda")
setStockcol(NULL)
setStockcol(paste0(getStockcol(), 70))
```
```{r showPDres}
pdRes
getMarkers(pdRes, fcol = "pd")
```

We can plot the results using the `plot2D` function.

```{r plotPDres, fig.width=9}
## Re-order the colours for the phenoDisco output
cl <- getMarkerClasses(pdRes, "pd", verbose = FALSE)
cols <- getStockcol()[seq(cl)]
ind <- grep("Pheno", cl, invert = TRUE)
cols[ind] <- getStockcol()[seq(cl)][1:length(ind)]
cols[-ind] <- getStockcol()[seq(cl)][(length(ind) + 1):length(cl)]

## Plot the input and output
par(mfrow = c(1, 2))
plot2D(pdRes, fcol = "phenodisco.input", 
       main = "phenoDisco input markers", col = getStockcol()[1:6])
addLegend(pdRes, fcol = "phenodisco.input", 
          where = "bottomleft", cex = .7)
plot2D(pdRes, fcol = "pd", main = "phenoDisco output", col = cols)
addLegend(pdRes, fcol = "pd", where = "bottomleft", cex = .7, col = cols)
```

The five new phenotype data clusters can be extracted and examined. 
In the code chunk below we write the results to a .csv file. We use 
the argument `fDataCols` to specify which columns of the `featureData` 
to write.

```{r writePDres, eval=FALSE}
fData(pdRes)$pd <- as.character(fData(pdRes)$pd)
write.exprs(pdRes, fDataCols = "pd", file = "pd-results.csv", sep = ",")
```

We can also examine the each phenotype intercatively and visualise 
their protein profiles by using the `pRolocVis` function in the 
`pRolocGUI`.

```{r visPDres, eval=FALSE}
pRolocVis(pdRes, fcol = "pd")
```
